---
title: "Storm"
author: "Oye Akinnikawe"
date: "December 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Impacts of Storms and other severe weather events in the US
## 1. Synopsis

This is project 2 of the Cousera Reproducible Research course. The goal of the 
project is to explore the NOAA storm database and analyze the impacts of events
types to population health and economic consequences.

The data covers events from the year 1950 to 2011. There are fewer recorded events
in earlier years compared to more recent years.

The analysis aims to investigate which different types of severe weather events are 
most harmful on the populations health in respect of general injuries and fatalities.
Further, the economic consequences will be analyzed by exploring the financial damage 
done to both general property and agriculture (i.e. crops)

## 2. Data Processing

The Storm data is a bzip2 file that was downloaded from the cousera website. 
The first step is to read the data.

```{r, data}
data <- read.csv("StormData.csv.bz2")
```

The required analysis deals with population health and economic analysis, therefore a subset of the required columns is created

```{r, data subset1}
storm <- data[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
                  "CROPDMG", "CROPDMGEXP")]
```

### 2.1 Population Health
Summarizing fatalities and injuries according to the event type.

```{r, data subset2}
library(plyr)
pop_affected <- ddply(storm, .(EVTYPE), summarize, fatalities = sum(FATALITIES),
                      injuries = sum(INJURIES))
fatal <- pop_affected[order(pop_affected$fatalities, decreasing = T), ]
injury <- pop_affected[order(pop_affected$injuries, decreasing = T), ]
```

### 2.2 Economic Cosequences
Since the exponential values are stored in a seperate column describing their 
value with letters (h = hundred, k = thousand, m = million, b = billion). 
A function that converts the letter value of the exponent to a usable number 
must be implemented.

```{r, function}
getExp <- function(e) {
    if (e %in% c("h", "H"))
        return(2)
    else if (e %in% c("k", "K"))
        return(3)
    else if (e %in% c("m", "M"))
        return(6)
    else if (e %in% c("b", "B"))
        return(9)
    else if (!is.na(as.numeric(e))) 
        return(as.numeric(e))
    else if (e %in% c("", "-", "?", "+"))
        return(0)
    else {
        stop("Invalid value.")
    }
}
```

Then, using this function, the proper values are calculated for property damage
and crop damage 

``` {r, Damage Calculation}
propExp <- sapply(storm$PROPDMGEXP, FUN=getExp)
storm$propDamage <- storm$PROPDMG * (10 ** propExp)
cropExp <- sapply(storm$CROPDMGEXP, FUN=getExp)
storm$cropDamage <- storm$CROPDMG * (10 ** cropExp)
```

Summarizing finacial damage for crops and property according to event type

``` {r, Economic Damage1}
econDamage <- ddply(storm, .(EVTYPE), summarize,propDamage = sum(propDamage), 
                    cropDamage = sum(cropDamage))
```

Omitting events that did not cause financial damage

``` {r, Economic Damage2}
econDamage <- econDamage[(econDamage$propDamage > 0 | econDamage$cropDamage > 0), ]
```

sorting the data in decreasing order

``` {r, Damage Sorted}
propDmgSorted <- econDamage[order(econDamage$propDamage, decreasing = T), ]
cropDmgSorted <- econDamage[order(econDamage$cropDamage, decreasing = T), ]
```

## 3. Results

### 3.1 Effects on population health

Lists of the Top 5 weather events affecting the populations health (injuries
 and deaths).

```{r, Event List}
head(injury[, c("EVTYPE", "injuries")],5)

head(fatal[, c("EVTYPE", "fatalities")],5)
```

Plotting the Top 5 population health events:

``` {r, Plots}
library(ggplot2)
library(grid)
library(gridExtra)

p1 <- ggplot(data=head(injury,5), aes(x=reorder(EVTYPE, injuries), y=injuries)) +
   geom_bar(fill="olivedrab",stat="identity")  + coord_flip() + 
    ylab("Total number of injuries") + xlab("Event type") +
    ggtitle("Health impact of weather events in the US - Top 5") +
    theme(legend.position="none")

p2 <- ggplot(data=head(fatal,5), aes(x=reorder(EVTYPE, fatalities), y=fatalities)) +
    geom_bar(fill="red4",stat="identity") + coord_flip() +
    ylab("Total number of fatalities") + xlab("Event type") +
    theme(legend.position="none")

grid.arrange(p1, p2, nrow =2) 
```

The plots shows that tornados are by far the most dangerous events when it comes to population health. 

###3.2. Economic Consequences

What are the top 5 weather events that have the greatest economic consequences?

```{r, Events Sorted2}
head(propDmgSorted[, c("EVTYPE", "propDamage")], 5)

head(cropDmgSorted[, c("EVTYPE", "cropDamage")], 5)
```

Plotting the Top 5 property damage and crop damage events: 

```{r, Plots2}
p1 <- ggplot(data=head(propDmgSorted,5), aes(x=reorder(EVTYPE, propDamage), 
          y=log10(propDamage), fill=propDamage )) +
    geom_bar(fill="darkred", stat="identity") + coord_flip() +
    xlab("Event type") + ylab("Property damage in dollars (log10)") +
    ggtitle("Economic impact of weather events in the US - Top 5") +
    theme(plot.title = element_text(hjust = 0))

p2 <- ggplot(data=head(cropDmgSorted,5), aes(x=reorder(EVTYPE, cropDamage), 
                                             y=cropDamage, fill=cropDamage)) +
    geom_bar(fill="goldenrod", stat="identity") + coord_flip() + 
    xlab("Event type") + ylab("Crop damage in dollars") + 
    theme(legend.position="none")

grid.arrange(p1, p2, ncol=1, nrow =2)
```
